Demographic Shifts in Toronto's Homeless Population: Evaluating and Reassessing City Solutions

### 2024-09-27 paper/paper.qmd
small changes for body
3rd edition of paper, with chat history updated

---
title: "Demographic Shifts in Toronto's Homeless Population: Evaluating and Reassessing City Solutions"
author: 
  - Junbo Li
thanks: "Code and data are available at: https://github.com/Junbo345/Homeless."
date: today
date-format: long
abstract: "Reducing the homeless population in the City of Toronto is an ongoing task, and every year the city government struggles to address the different needs of homeless populations. In our paper, we analyze the demographic change of the homeless population in the past six years with data from Toronto's shelter system. We show that the proportion of chronic, refugees, and working-age all rose in the past six years. These findings indicate the city's current solution which focuses on building houses is no longer effective. We further propose new policies like improving public health systems to beat the latest crisis."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Homelessness is defined as "the situation of an individual, family or community without stable, safe, permanent, appropriate housing, or the immediate prospect, means and ability of acquiring it" [@canadianobservatory2012definition]. According to @gaetz2014state, approximately 23,500 Canadian people experience every type of homelessness every year. On SEP 12, 2024, there are still more than 9500 people staying in shelters in the city of Toronto [@toronto_shelter_census]. To address this ongoing crisis, the Toronto city government launched the supportive housing development plan, which aims to build more affordable housing for homeless people [@toronto2024affordable]. A recent two-year plan starting in 2022 aims to build 4,000 more affordable homes as well as supportive housing by the end of 2024 [@toronto2024affordable].

While the city of Toronto has concentrated its efforts on building more affordable housing, our analysis in this paper has found that the underlying demographics of homeless populations changed significantly. This would indicate that the traditional housing-centric plan would no longer solve the newly evolved homeless problem.

In this paper, we delve into the demographic shift of Toronto's homeless population in greater detail by studying the monthly data from Toronto's shelter system from 2018 Jan to 2024 Jun [@toronto_shelter_data]. We start by outlining the data source, the way we accessed the data, as well as procedures we used to clean and prepare the data in analysis in **Data** section. Following this, in **Result** section, we draw the bar graph demonstrating the trend in demographic changes and present our key findings. Finally, in **Discussion** section, we analyze the implications of the trends we found and offer recommendations for policy changes that would address the needs of different demographic groups.

@@

We dropped the only missing row corresponding to July 2024. We then selected our columns of interest, namely, the total number of homeless people every month (**actively_homeless**), as well as **refugee** and **chronic status** in the column population. In addition, we categorized age groups based on their suitability for work: transforming ages under 24 for the **early career stage**, ages between 25 and 54 for **working age**, and ages above 55 for **pre-retirement and retirement age**. The classification of ages 25-54 as working age is aligned with the Canadian government's survey, which defines individuals aged 25-54 as core-aged workers [@statcan2024]. Finally, we replaced the monthly data with averaged quarterly data for better analysis and visualization over a prolonged period. A sample of cleaned data can be seen in @tbl-cleandata, and summary statistics in @tbl-summary.

@@

## Results

>>>>

---
title: "Demographic Shifts in Toronto's Homeless Population: Evaluating and Reassessing City Solutions"
author: 
  - Junbo Li
thanks: "Code and data are available at: https://github.com/Junbo345/Homeless."
date: today
date-format: long
abstract: "Homelessness in the City of Toronto has been a growing issue, with the city government facing difficulties in meeting the needs of different groups within the homeless population. In this paper, we analyze changes in the demographics of Toronto’s homeless population over the past six years, using data from the city's shelter system. We found that the proportion of individuals experiencing chronic homelessness, refugees, and working-age adults has increased. These findings suggest that the city's current focus on building housing is not enough, and we propose improvements in public health services to better address the evolving crisis."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Homelessness is defined as "the situation of an individual, family, or community without stable, safe, permanent, appropriate housing, or the immediate prospect, means, and ability of acquiring it" [@canadianobservatory2012definition]. In Canada, around 23,500 people experience various forms of homelessness every year [@gaetz2014state], with more than 9,500 individuals currently living in shelters in Toronto as of September 12, 2024 [@toronto_shelter_census]. To address this crisis, the Toronto city government has implemented several initiatives, including a supportive housing development plan, which aims to build more affordable housing for homeless populations [@toronto2024affordable]. A key part of this plan is the 2022-2024 two-year effort to construct 4,000 new affordable homes [@toronto2024affordable].

While the city has made progress in expanding housing, we find that a housing-centric approach may not be sufficient to address the needs of the evolving homeless population. Our analysis reveals significant demographic changes in Toronto’s homeless population over the past six years, including increases in the proportion of chronically homeless individuals, refugees, and working-age adults. These demographic shifts suggest that the traditional emphasis on housing development alone is no longer be an adequate solution.

In this paper, we explore these demographic changes using monthly data from Toronto's Shelter Management Information System (SMIS) from January 2018 to June 2024 [@toronto_shelter_data]. We detail the data collection process, cleaning methods, and key findings in the Data section. In the final section, Discussion, we assess the implications of our results for policy and propose recommendations for more comprehensive strategies that extend beyond housing solutions to address the diverse needs of Toronto's homeless population.

@@

We dropped the only missing row corresponding to July 2024. We then selected our columns of interest, namely, the total number of homeless people every month (**actively_homeless**), as well as **refugee** and **chronic status** in the column population. In addition, we categorized age groups based on their suitability for work: transforming ages under 24 for the **early career stage**, ages between 25 and 54 for **working age**, and ages above 55 for **pre-retirement and retirement age**. The classification of ages 25-54 as working age is aligned with the Canadian government's survey, which defines individuals aged 25-54 as core-aged workers [@statcan2024]. Finally, we replaced the monthly data with averaged quarterly data for better analysis and visualization over a prolonged period. A sample of cleaned data can be seen in @tbl-cleandata, and the key features for each variable is in @tbl-summary.

@@

## Demographic trend with respect to time
